### 2014-12-16 text/Chapter05.tex
Spell checking

In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware prorgamming
languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
and we introduce the \emph{structural coeffect calculus}.

@@

Our work follows the ``language semantics'' style in that we provide a structural semantics to
the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
the meta-langauge style, which extends the terms and types of a language with constructs for explicitly
manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
annotations as stand-alone types (indexed comonads only appear in the semantics).

@@

Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma 
(Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
for all our examples; it is needed for the structural version of the subtitution lemma 
(Lemma~\ref{thm:structural-substitution}).

@@

The type system for the structural coeffect calculus is similar to sub-structural type systems
in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
weakening, exchange or contraction -- this is done by checking the types of sub-expressions
in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
perform corresponding transformation on the coeffect annoation.

@@

Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
in the context, performing the same reordering on the coeffect vector. Finally recall that 
variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
allows re-using a variable. We can type check sub-expressions using two separate varaibles and 
then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
is the only place in the structural coeffect system where context requiremens are combined
(semantically, this is where the available context is shared).

@@

It is important to note that this gives us a systems with \emph{different} properties. The 
information are now tracked per-variable rather than for enitre contexts. For data-flow,
we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
for a pointwise extension of the $+$ operator, such that
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
%
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
The rule appears similar to the flat one, but there is an important difference. Because of the
structural nature of the type system, it only increments the required number of values for 
variables that are used in the expression $e$. Annotations of other variables can be left 
unchanged.

@@

The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This assymmetry
is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
takes a vector of variables, but always produces just a single value. Thus the resulting function 
also takes a vector of variables, but always returns a context with a vector containing just one value.
In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
no way of constructing values containing $\xtimes$.

@@

\subsection{Semantics of structural caluculus}

@@

\paragraph{Expressions.}
A semantics of strucutural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
the previous chapter, the semantics is written in a programming language style using constructs
such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
can equally be written using primitives such as currying, uncurrying, function pairing etc.

@@

The transformations on the middle part are quite simple. The (\emph{sub}) rule 
uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a varaible while 
splitting its additional context.

@@

\noindent
As in the flat calculus, the primitive operations of the structural indexed comonad are each
annotated with a unique operations provided by the coeffect annotations. This means that the
semantics fully determines the structure of the typing rules of the strucutral coeffect
calculus.

@@

The $\ident{dup}_{m,n}$ operation needs to produce a two-varaible context containing $m$ and $n$ values,
respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable
and \ident{lift} drops additional past values that are not required.

@@

The chapter presents two technical results. The first is the semantics of structural coeffect
calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
it unifies our three examples (bounded reuse, data-flow and liveness) and it demonstrates
suitability of the type system. The second technical result is equational theory for the structural
coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
typing preservation property for any instance of the structural coeffect calculus. These two 
strong properties are desirable for programming languages, but are often not satisfied 
(\eg~by langauges with effects).

>>>>

In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware programming
languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
and we introduce the \emph{structural coeffect calculus}.

@@

Our work follows the ``language semantics'' style in that we provide a structural semantics to
the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
the meta-language style, which extends the terms and types of a language with constructs for explicitly
manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
annotations as stand-alone types (indexed comonads only appear in the semantics).

@@

Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma 
(Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
for all our examples; it is needed for the structural version of the substitution lemma 
(Lemma~\ref{thm:structural-substitution}).

@@

The type system for the structural coeffect calculus is similar to sub-structural type systems
in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
weakening, exchange or contraction -- this is done by checking the types of sub-expressions
in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
perform corresponding transformation on the coeffect annotation.

@@

Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
in the context, performing the same reordering on the coeffect vector. Finally recall that 
variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
allows re-using a variable. We can type check sub-expressions using two separate variables and 
then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
is the only place in the structural coeffect system where context requirements are combined
(semantically, this is where the available context is shared).

@@

It is important to note that this gives us a systems with \emph{different} properties. The 
information are now tracked per-variable rather than for entire contexts. For data-flow,
we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
for a pointwise extension of the $+$ operator, such that
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
%
\begin{equation*}
\tyrule{prev}
  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
\end{equation*}
%
The rule appears similar to the flat one, but there is an important difference. Because of the
structural nature of the type system, it only increments the required number of values for 
variables that are used in the expression $e$. Annotations of other variables can be left 
unchanged.

@@

The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
takes a vector of variables, but always produces just a single value. Thus the resulting function 
also takes a vector of variables, but always returns a context with a vector containing just one value.
In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
no way of constructing values containing $\xtimes$.

@@

\subsection{Semantics of structural calculus}

@@

\paragraph{Expressions.}
A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
the previous chapter, the semantics is written in a programming language style using constructs
such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
can equally be written using primitives such as currying, uncurrying, function pairing etc.

@@

The transformations on the middle part are quite simple. The (\emph{sub}) rule 
uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a variable while 
splitting its additional context.

@@

\noindent
As in the flat calculus, the primitive operations of the structural indexed comonad are each
annotated with a unique operations provided by the coeffect annotations. This means that the
semantics fully determines the structure of the typing rules of the structural coeffect
calculus.

@@

The $\ident{dup}_{m,n}$ operation needs to produce a two-variable context containing $m$ and $n$ values,
respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable
and \ident{lift} drops additional past values that are not required.

@@

The chapter presents two technical results. The first is the semantics of structural coeffect
calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
it unifies our three examples (bounded reuse, data-flow and liveness) and it demonstrates
suitability of the type system. The second technical result is equational theory for the structural
coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
typing preservation property for any instance of the structural coeffect calculus. These two 
strong properties are desirable for programming languages, but are often not satisfied 
(\eg~by languages with effects).
